Towards a Verified Tableau Prover for a Ground Fragment of Set Theory

### 2023-02-14 paper/CADE/paper.tex
Continue with paper

@@

@@

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  \keywords{First keyword \and Second keyword}
\end{abstract}

@@

\subsection{Contributions}
In this paper, we present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation follows the paper closely but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

@@

While a lot of work has gone into the verification of decision procedures, seemingly little has been put towards the verification of tableau calculi.
In Lean, there is a verified tableau calculus for basic modal logic~\cite{modal_logic_lean}.
In Coq, there is a formalisation of a tableau calculus for ecumenical logic~\cite{ecumenical_logic}; it is, however, missing a completeness proof.
Concerning Isabelle/HOL, there is formalisation of a sound and complete tableau calculus for hybrid logic~\cite{hybrid_logic} available in the \textit{Archive of Formal Proofs}~\cite{hybrid_logic_afp}.
The termination proof is part of ongoing work~\cite{hybrid_logic_workshop}.
% TODO: Formalisation de la logique temporelle dans le coq.

@@

Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
The cardinality of a set \lstinline!A! is denoted by \lstinline!card A! and the image of \lstinline!A! under \lstinline!f! by \lstinline!f ` A!.

@@

Lastly, we consider the subformulae of a formula and define a function that computes them.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) = {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) = {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}
\textit{Literals} are a type of subformulae of particular importance, so we define a predicate for them.
\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True | is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}
\subsection{Semantics}
We base the semantics of \MLSS{} on the von Neumann hierarchy $\mathcal{V}$ of sets that is defined inductively as
  \[
    \begin{array}{rclr}
      \mathcal{V}_0 & = & \emptyset, \\
      \mathcal{V}_{\alpha + 1} & = & \mathcal{P}(\mathcal{V}_\alpha) & \text{for each ordinal $\alpha$}, \\
      \mathcal{V}_{\lambda} & = & \bigcup_{\mu < \lambda} \mathcal{V}_\mu & \text{for each limit ordinal $\lambda$}, \\
      \mathcal{V} & = & \bigcup_\alpha \mathcal{V}_\alpha, \\
    \end{array}
  \]
where $\mathcal{P}(S)$ is the powerset of $S$.
The sets in $\mathcal{V}$ are well-founded, i.e.\ there can be no membership cycle in $\mathcal{V}$.

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitation is provided by \citeauthor{zfc_in_hol_afp} in an entry~\cite{zfc_in_hol_afp} of the \textit{Archive of Formal Proofs}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
Two articles~\cite{zfc_in_hol1,zfc_in_hol2} have been written that provide further context to this entry.
The entry declares a type \lstinline!V! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!elts :: V $\Rightarrow$ V set! maps a set to its elements.
      Note that even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class (relative to \lstinline!V!).
      This class only constitutes a set (relative to \lstinline!V!) if the class is small.
  \item The predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, which is polymorphic in \lstinline!'a!, indicates whether a class is small.
  \item The function \lstinline!vset :: 'a set $\Rightarrow$ V! turns a class into a set of type \lstinline!V! if it is small. 
\item The usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: V!.
\end{itemize}

Equipped with the above, we define the interpretation function \lstinline!&\Ist&! for set terms that interprets a set term with respect to a valuation function \lstinline!M :: 'a $\Rightarrow$ V! for variables.
\begin{lstlisting}
fun &\Ist& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  &\Ist& M $\emptyset$ = 0
| &\Ist& M (Var x) = M x
| &\Ist& M (Single s) = vset {&\Ist& M s}
| &\Ist& M (s $\unionS$ t) = &\Ist& M s $\sqcup$ &\Ist& M t 
| &\Ist& M (s $\interS$ t) = &\Ist& M s $\sqcap$ &\Ist& M t 
| &\Ist& M (s $\diffS$ t) = &\Ist& M s $-$ &\Ist& M t 
\end{lstlisting}
The interpretation function \lstinline!I$_\text{sa}$! for set atoms is straightforward as well.
\begin{lstlisting}
fun &\Isa& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ bool where
  &\Isa& M (s $\inS$ t) $\longleftrightarrow$ &\Ist& M s $\in$ elts (&\Ist& M t)
| &\Isa& M (s $\eqS$ t) $\longleftrightarrow$ &\Ist& M s = &\Ist& M t
\end{lstlisting}

@@

\section{Completeness of the Calculus\label{sec:complete}}
For completeness of the calculus, we need to show that every unsatisfiable formula has a closed tableau or, conversely, that the formula is satisfiable if there is a saturated and open branch in the tableau.
To facilitate inductive reasoning, we show a stronger statement by constructing a model \lstinline!M! such that \lstinline!M $\models$ $\phi$! for all \lstinline!$\phi$ $\in$ set b!.
At the core of the model, there is a \textit{realisation} function that maps set terms to sets of type \lstinline!V!.
A subset of the witnesses, which we call \textit{pure} witnesses, receives special treatment from the realisation function for reasons that will become apparent in \autoref{sec:pwits}.
The collection of set terms of a branch can thus be partitioned into two collections as defined below.
\begin{lstlisting}
definition pwits :: 'a branch $\Rightarrow$ 'a set where
  pwits b $\equiv$ {c $\in$ wits b. $\forall$t $\in$ subterms (last b).
                          AT (Var c $\eqS$ t) $\notin$ set b $\land$
                          AT (t $\eqS$ Var c) $\notin$ set b}

@@

We aim to construct a syntactic model that is derived from the membership literals \lstinline!s $\inS$ t! in the branch.
To this end, we construct a graph whose vertices are the disjoint union of the sets above and there is an edge from \lstinline!s! to \lstinline!t! in the graph if, and only if, \lstinline!s $\inS$ t! is in \lstinline!b!.
Note that we use \citeauthor{graph_theory_afp}'s graph library~\cite{graph_theory_afp} that represents a graph as a record of vertices, arcs (directed edges), and two functions \lstinline!tail! and \lstinline!head! that map an arc to its source, respectively target, vertex.
\begin{lstlisting}
definition bgraph b $\equiv$
  let vs = Var ` pwits b $\cup$ subterms' b
  in $\llparenthesis$ verts = vs, arcs = {(s, t). (s $\inS$ t) $\in$ set b},
     &$\phantom{\llparenthesis}$& tail = fst, head = snd $\rrparenthesis$
\end{lstlisting}

The realisation function is defined relative to this graph.
As mentioned before, the pure witnesses \lstinline!Var ` pwits b! and the rest of the set terms \lstinline!subterms' b! are treated differently by the realisation function.
Terms in the latter set are evaluated in accordance to the structure of the graph, i.e.\ the realisation of a vertex is defined as the union of the realisations of the parent vertices. 
For the former set, we choose a function \lstinline!I! that assigns the witnesses in \lstinline!Var ` pwits b! pairwise distinct sets with cardinality greater than that of the vertices.
We can always choose such a function since we assume an infinite universe of variables.
Then, we return the singleton set \lstinline!vset {I x}!, which, together with the cardinality constraint, guarantees that realisations are distinct between pure witnesses themselves as well as between pure witnesses and set terms.  
The notation \lstinline!u $\rightarrow_\text{G}$ s! in the definition below indicates that there is an edge from \lstinline!u! to \lstinline!s! in the graph \lstinline!G!.
\begin{lstlisting}
abbreviation parents G s $\equiv$ {u. u $\rightarrow_\text{G}$ s}

function realise :: 'a pset_term $\Rightarrow$ V where
  x $\in$ Var ` pwits b $\Longrightarrow$ realise x = vset {I x}
| x $\in$ subterms' b
  $\Longrightarrow$ realise t = vset {realise ` parents (bgraph b) s}
| x $\notin$ verts G $\Longrightarrow$ realise x = 0
\end{lstlisting}

Again, we need to ensure that the assumptions of the equations are non-overlapping and that the function terminates.
The former is taken care of by automation, leaving us to prove termination.
The assumption that \lstinline!b! is open implies that there are no membership cycles and thus \lstinline!bgraph b! is acyclic.
Now, consider the ancestors \lstinline!av! of a vertex \lstinline!v!, which are those vertices from which we can reach \lstinline!v!.
Due to acyclicity, it holds that the ancestors \lstinline!ap! of a parent \lstinline!p! of \lstinline!v! are a subset of \lstinline!av!.
They are also a proper subset of \lstinline!av! since \lstinline!p! is an ancestor of \lstinline!v! but not of \lstinline!p! itself.
Together with the fact that \lstinline!bgraph b! is finite, it follows that the cardinality of the ancestors decreases in each recursive call, thus proving the termination of \lstinline!realise!.

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b and c $\in$ pwits b
  shows (Var c $\eqS$ t) $\notin$ set b and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: V!, were they not treated separately. 
To see that this is a problem, consider the branch
\begin{lstlisting}
  b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]
\end{lstlisting}
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y,
   Var x $\inS$ Var s, Var x $\notinS$ Var t,
   Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

The proof of \hyperref[lst:lemma_2]{\lstinline!lemma_2!} is more technical than interesting so we refer the reader to the formalisation.

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realisation function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realisation function models all literals in the branch; more formally, the following propositions hold:
\begin{enumerate}[label={(\roman*)}]
  \item\label{it:at_mem} We have \lstinline!realise s $\in$ elts (realise t)! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realise s $=$ realise t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realise s $\neq$ realise t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realise s $\notin$ elts (realise t)! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \hyperref[lst:lemma_2]{\lstinline!lemma_2!}, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle, so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ pwits b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \hyperref[lst:lemma_2]{\lstinline!lemma_2!}.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\in$ subterms' b! must hold.
  Now, assume for contradiction that
\begin{lstlisting}
  realise s $\neq$ realise t
\end{lstlisting}
  which implies that
\begin{lstlisting}
  elts (realise s) $\neq$ elts (realise t).
\end{lstlisting}
  Without loss of generality --- the other case is symmetric --- we obtain an \lstinline!e! such that
\begin{lstlisting}
  e $\in$ elts (realise s) $\land$ e $\notin$ elts (realise t). 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realise!, we obtain an \lstinline!u! where \lstinline!e = realise u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ set b! and the saturation of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realise u $\in$ elts (realise t) = e $\in$ elts (realise t)
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!e $\notin$ elts (realise t)!.
\end{proof}

\noindent The results on literals can now be lowered to set terms.
\begin{enumerate}[label=(\alph*)]
  \item\label{it:empty} It holds that \lstinline!realise $\emptyset$ = 0!.
  \item\label{it:op} Let \lstinline!$\star_\text{s}$ $\in$ {$\unionS$, $\diffS$, $\interS$}!. If the term \lstinline!s $\star_\text{s}$ t! occurs in \lstinline!subterms b!, then
    \begin{center}
    \lstinline!realise (s $\star_\text{s}$ t) = realise s $\star$ realise t!.
    \end{center}
  \item\label{it:single} If \lstinline!Single t $\in$ subterms b!, then
    \begin{center}
    \lstinline!realise (Single t) = vset {realise t}!.
    \end{center}
\end{enumerate}
\begin{proof}[Proof of Proposition~\ref{it:empty}]
  By definition, \lstinline!realise $\emptyset$! is equal to \lstinline!0! if \lstinline!$\emptyset$! is not a vertex of \lstinline!bgraph b!.
  Otherwise, \lstinline!$\emptyset$! must be in \lstinline!subterms' b! as it is not a witness.
  Moreover, since \lstinline!b! is open, we know \lstinline!s $\inS$ $\emptyset$! is not in \lstinline!b! for any \lstinline!s!, implying that the term \lstinline!$\emptyset$! has no parents in \lstinline!bgraph b!.
  Ultimately, we obtain \lstinline!realise $\emptyset$ = 0!.
\end{proof}
\begin{proof}[Proof of Proposition~\ref{it:op} for $\star_\texttt{s} = -_\texttt{s}$]
  For equality, it suffices to prove subset inclusion in both directions.

  \textbf{Direction $\subseteq$. } Fix some \lstinline!e! in \lstinline!elts (realise (s $\diffS$ t))!.
  Again, we obtain an \lstinline!u! with 
\begin{lstlisting}
  e = realise u $\land$ u $\rightarrow_\text{bgraph b}$ (s $\diffS$ t).
\end{lstlisting}
  We therefore have \lstinline!(u $\inS$ s $\diffS$ t) $\in$ set b! which, due to \lstinline!b! being saturated, implies that \lstinline!u $\inS$ s! is in \lstinline!b!.
  Furthermore, the fifth branching rule (see \autoref{tab:bexpands}) must have its subsumption condition met due to saturation of \lstinline!b!, meaning that either one of \lstinline!u $\inS$ t! or \lstinline!u $\notinS$ t! must be in the branch.
  Only the latter, however, can be the case since the branch is open.
  By applying Proposition~\ref{it:at_mem} and~\ref{it:af_mem} we obtain
\begin{lstlisting}
  realise u $\in$ realise s $\land$ realise u $\notin$ realise t
\end{lstlisting}
  and together with \lstinline!e = realise u! we arrive at the goal 
\begin{lstlisting}
  e $\in$ elts (realise s - realise t).
\end{lstlisting}

\textbf{Direction $\supseteq$. } Let \lstinline[columns=fullflexible]!e $\in$ elts (realise s - realise t)! for some fixed \lstinline!e!.
  As before, we obtain an \lstinline!u! such that \lstinline!e! is the realisation of \lstinline!u!, i.e.\ \lstinline!e = realisation u!, and
\begin{lstlisting}
  u $\rightarrow_\text{bgraph b}$ s $\land$ u $\nrightarrow_\text{bgraph b}$ t.
\end{lstlisting}
  An immediate consequence is that \lstinline!u $\inS$ s! is in \lstinline!b!.
  Moreover, we claim that \lstinline!u $\notinS$ t! is also in \lstinline!b!.
  To see that this is true, recall that \lstinline!b! is saturated and thus the subsumption condition of the fifth branching rule must hold.
  This implies that either \lstinline!u $\inS$ t! or \lstinline!u $\notinS$ t! is in \lstinline!b!.
  But the latter being in \lstinline!b! would be a contradiction to \lstinline!u $\nrightarrow_\text{bgraph b}$ t! meaning that the former must hold.
  Ultimately, we infer that \lstinline!u $\in$ s $\diffS$ t! is in \lstinline!b! by invoking saturation again.
  The last step is to apply Proposition~\ref{it:at_mem} in order to obtain the goal
  \begin{lstlisting}
    realise u $\in$ elts (realise (s $\diffS$ t))
          = e $\in$ elts (realise (s $\diffS$ t)).
  \end{lstlisting}
\end{proof}

>>>>

\usepackage{appendix}

@@

\newcommand{\hfmem}{\boldsymbol{\in}}
\newcommand{\hfnotmem}{\boldsymbol{\notin}}

@@

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  \keywords{First TODO \and Second TODO}
\end{abstract}

@@

\subsection{Contributions}
In this paper, we present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation follows the paper closely but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
  \item We restrict one branching rule TODO.
\end{itemize}

@@

While a lot of work has gone into the verification of decision procedures, seemingly little has been put towards the verification of tableau calculi.
In Lean, there is a verified tableau calculus for basic modal logic~\cite{modal_logic_lean}.
In Coq, there is a formalisation of a tableau calculus for ecumenical logic~\cite{ecumenical_logic}; it is, however, missing a completeness proof.
Concerning Isabelle/HOL, there is formalisation of a sound and complete tableau calculus for hybrid logic~\cite{hybrid_logic} available in the \textit{Archive of Formal Proofs}~\cite{hybrid_logic_afp}.
The termination proof is part of ongoing work~\cite{hybrid_logic_workshop}.
TODO: Mention SteP, Check reviewer comment.
% TODO: Formalisation de la logique temporelle dans le coq.

@@

Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
The cardinality of a set \lstinline!A! is denoted by \lstinline!|A|! and the image of \lstinline!A! under \lstinline!f! by \lstinline!f ` A!.

@@

Similarly to \lstinline!vars!, we introduce the overloaded constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!, which returns all the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The definition of \lstinline!subterms! and \lstinline!subfms! is standard so we omit them here and refer to the appendix.

\textit{Literals} are a type of subformulae of particular importance, so we define a predicate for them.
\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True | is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}

\subsection{Semantics}
The original paper~\cite{new_fast_tableau} bases the semantics of \MLSS{} on the von Neumann hierarchy of sets $\mathcal{V}$. 
While there is an entry~\cite{zfc_in_hol_afp} in the \textit{Archive of Formal Proofs} (AFP) that axiomatises this hierarchy,
we instead use the hierarchy of \textit{hereditarily finite sets} (HF sets) which fulfil all the same axioms as $\mathcal{V}$ except for the axiom of infinity.
The HF sets are sufficient for our purposes as \MLSS{} is not powerful enough to force infinite models~\cite[Chapter~14.2]{set_theory}.
In contrast to $\mathcal{V}$, the HF sets are directly representable in Isabelle/HOL, and indeed, there is an AFP entry~\cite{hf_AFP} that formalises them.
The entry defines a type \lstinline!hf! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!HF :: hf set $\Rightarrow$ set! that converts a finite set of HF sets into an HF set.
\item The usual set operations such as equality ($=$), membership ($\hfmem$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: hf!.
\end{itemize}

Equipped with the above, we define the interpretation functions
\begin{lstlisting}
  &\Ist& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_term $\Rightarrow$ hf &\textnormal{and}&
  &\Isa& :: ('a $\Rightarrow$ hf) $\Rightarrow$ 'a pset_atom $\Rightarrow$ hf
\end{lstlisting}
in the standard way, i.e.\ by mapping each syntactic construct to the corresponding operation on HF sets and interpreting variables with respect to a given valuation function \lstinline!M :: 'a $\Rightarrow$ hf!.
For the concrete definition we refer to the appendix.

@@

\section{Completeness of the Calculus\label{sec:complete}}
For completeness of the calculus, we need to show that every unsatisfiable formula has a closed tableau or, conversely, that the formula is satisfiable if there is a saturated and open branch in the tableau.
To facilitate inductive reasoning, we show a stronger statement by constructing a model \lstinline!M! such that \lstinline!M $\models$ $\phi$! for all \lstinline!$\phi$ $\in$ set b!.
At the core of the model, there is a \textit{realisation} function that maps set terms to sets of type \lstinline!hf!.
A subset of the witnesses, which we call \textit{pure} witnesses, receives special treatment from the realisation function for reasons that will become apparent in \autoref{sec:pwits}.
The collection of set terms of a branch can thus be partitioned into two collections as defined below.
\begin{lstlisting}
definition pwits :: 'a branch $\Rightarrow$ 'a set where
  pwits b $\equiv$ {c $\in$ wits b. $\forall$t $\in$ subterms (last b).
                          AT (Var c $\eqS$ t) $\notin$ set b $\land$
                          AT (t $\eqS$ Var c) $\notin$ set b}

@@

We aim to construct a syntactic model that is derived from the membership literals \lstinline!s $\inS$ t! in the branch.
To this end, we construct a graph whose vertices are the disjoint union of the sets above and there is an edge from \lstinline!s! to \lstinline!t! in the graph if, and only if, \lstinline!s $\inS$ t! is in \lstinline!b!.
Note that we use \citeauthor{graph_theory_afp}'s graph library~\cite{graph_theory_afp} that represents a graph as a record of vertices, arcs (directed edges), and two functions \lstinline!tail! and \lstinline!head! that map an arc to its source and target vertex, respectively.
\begin{lstlisting}
definition bgraph b $\equiv$
  let vs = Var ` pwits b $\cup$ subterms' b
  in $\llparenthesis$ verts = vs, arcs = {(s, t). (s $\inS$ t) $\in$ set b},
     &$\phantom{\llparenthesis}$& tail = fst, head = snd $\rrparenthesis$
\end{lstlisting}

The realisation function is defined relative to this graph.
As mentioned before, the pure witnesses \lstinline!Var ` pwits b! and the rest of the set terms \lstinline!subterms' b! are treated differently by the realisation function.
Terms in the latter set are evaluated in accordance to the structure of the graph, i.e.\ the realisation of a vertex is defined as the union of the realisations of the parent vertices. 
For the former set, we choose a function \lstinline!I! that assigns the witnesses in \lstinline!Var ` pwits b! pairwise distinct sets with cardinality greater than that of the vertices.
We can always choose such a function since we assume an infinite universe of variables.
Then, we return the singleton set \lstinline!HF {I x}!, which, together with the cardinality constraint, guarantees that realisations are distinct between pure witnesses themselves as well as between pure witnesses and set terms.  
The notation \lstinline!u $\rightarrow_\text{G}$ s! in the definition below indicates that there is an edge from \lstinline!u! to \lstinline!s! in the graph \lstinline!G!.
\begin{lstlisting}
abbreviation parents G s $\equiv$ {u. u $\rightarrow_\text{G}$ s}

function realise :: 'a pset_term $\Rightarrow$ V where
  x $\in$ Var ` pwits b $\Longrightarrow$ realise x = HF {I x}
| x $\in$ subterms' b
  $\Longrightarrow$ realise t = HF {realise ` parents (bgraph b) s}
| x $\notin$ verts G $\Longrightarrow$ realise x = 0
\end{lstlisting}

Again, we need to ensure that the assumptions of the equations are non-overlapping and that the function terminates.
The former is taken care of by automation, leaving us to prove termination.
The assumption that \lstinline!b! is open implies that there are no membership cycles and thus \lstinline!bgraph b! is acyclic.
TODO: Shorten this proof.
Now, consider the ancestors \lstinline!av! of a vertex \lstinline!v!, which are those vertices from which we can reach \lstinline!v!.
Due to acyclicity, it holds that the ancestors \lstinline!ap! of a parent \lstinline!p! of \lstinline!v! are a subset of \lstinline!av!.
They are also a proper subset of \lstinline!av! since \lstinline!p! is an ancestor of \lstinline!v! but not of \lstinline!p! itself.
Together with the fact that \lstinline!bgraph b! is finite, it follows that the cardinality of the ancestors decreases in each recursive call, thus proving the termination of \lstinline!realise!.

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b and c $\in$ pwits b
  shows (Var c $\eqS$ t) $\notin$ set b and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: hf!, were they not treated separately. 
To see that this is a problem, consider the branch
\begin{lstlisting}
  b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]
\end{lstlisting}
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y,
   Var x $\inS$ Var s, Var x $\notinS$ Var t,
   Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

The proof of \hyperref[lst:lemma_2]{\lstinline!lemma_2!} is more technical than interesting so we refer the reader to the formalisation or paper TODO.

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realisation function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realisation function models all literals in the branch; more formally, the following propositions hold:
\begin{enumerate}[label={(\roman*)}]
  \item\label{it:at_mem} We have \lstinline!realise s $\hfmem$ realise t! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realise s $=$ realise t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realise s $\neq$ realise t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realise s $\hfnotmem$ realise t! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \hyperref[lst:lemma_2]{\lstinline!lemma_2!}, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle, so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ pwits b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \hyperref[lst:lemma_2]{\lstinline!lemma_2!}.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\in$ subterms' b! must hold.
  Now, assume for contradiction that
\begin{lstlisting}
  realise s $\neq$ realise t.
\end{lstlisting}
  Without loss of generality --- the other case is symmetric --- we obtain an \lstinline!e! such that
\begin{lstlisting}
  e $\hfmem$ realise s $\land$ e $\hfnotmem$ realise t. 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realise!, we obtain an \lstinline!u! where \lstinline!e = realise u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ set b! and the saturation of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realise u $\hfmem$ realise t = e $\hfmem$ realise t
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!e $\hfnotmem$ realise t!.
\end{proof}

\noindent The results on literals can now be lowered to set terms.
All of the proofs are straightforward so we refer the reader to the formalisation.
\begin{enumerate}[label=(\alph*)]
  \item\label{it:empty} It holds that \lstinline!realise $\emptyset$ = 0!.
  \item\label{it:op} Let \lstinline!$\star_\text{s}$ $\in$ {$\unionS$, $\diffS$, $\interS$}!. If the term \lstinline!s $\star_\text{s}$ t! occurs in \lstinline!subterms b!, then
    \begin{center}
    \lstinline!realise (s $\star_\text{s}$ t) = realise s $\star$ realise t!.
    \end{center}
  \item\label{it:single} If \lstinline!Single t $\in$ subterms b!, then
    \begin{center}
    \lstinline!realise (Single t) = HF {realise t}!.
    \end{center}
\end{enumerate}

The final step to obtain a proper model is to connect the realisation function to the semantics as defined in \autoref{sec:semantics}.
For set terms, we can use the Propositions~\ref{it:empty}--\ref{it:single} to prove the lemma below by induction on \lstinline!t!.
\begin{lstlisting}
lemma
  assumes t $\in$ subterms b
  shows &\Ist& ($\lambda$x. realise (Var x)) t = realise t
\end{lstlisting}
Lifting the above result to formulae yields the coherence of \lstinline!b! as the original paper~\cite{new_fast_tableau} calls it.
The proof is a tedious but straightforward induction on the the size of the formulae.
\begin{lstlisting}[label={lst:coherence}]
lemma coherence:
  assumes $\phi$ $\in$ set b shows ($\lambda$x. realise (Var x)) $\models$ $\phi$
\end{lstlisting}
The coherence property finishes the proof of completeness of the calculus as it gives us a model for every formula in an open and saturated branch.
